Reasoning Graphically about Quantum Computation

### 2008-03-03 papers/reasoning-graphically/paper.tex
intro rewritten
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1396 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

Recent work in quantum computation has emphasised the use of graphical
languages motivated by the underlying logical structure of quantum
mechanics itself.  These techniques have a number of advantages over
the convential matrix-based approach to quantum mechanics, but in
particular the graph can directly represent the structure of quantum
states explicitly, including the

@@

Can formalise quantum mechanics in an abstract setting using compact
or $\dag$-compact categories.

We will give a description of graphs in general;  a specialisation of
this gives the structure required for compact closed  categories.  The
graph structure suffices to give a tight representation.

We then introduce a concrete set of generators and equations for use
in quantum computation.  We need rewrites to capture the non-logical
axioms of this structure.   It turns out that we can most easily express
these equations using an infinite family of ``spider rules'';  this
motivates the  more general notion of graph patterns.

@@

The intuition behind this definition is that the subgraph of $G$
determined by $V'$ should be preserved exactly by $f$; it should
contain $G$ without any extra incident egdes into the verticies $V'$
whereas other vertices may be identified and may contain additional
verticies.

@@

\begin{theorem}
  $G \leq_v H \Leftrightarrow \vinterp{G} \superseteq \vinterp{H}$. 
\end{theorem}

>>>>

Recent work in quantum computation has emphasised the use of graphical
languages motivated by the underlying logical structure of quantum
mechanics
itself~\cite{AbrCoe:CatSemQuant:2004,Selinger:dagger:2005,Coecke2005Kindergarten-Qu,Coecke2006POVMs-and-Naima,Coecke2006Quantum-Measure}.
These techniques have a number of advantages over the convential
matrix-based approach to quantum mechanics:

\begin{itemize}
\item The visual representation abstracts over the values in the
  matricies. This removes detail that requires a lot of work for a
  human to interpret.

\item Many properties have a natural graphical representation. For
  example, non entanglement is can be infered from disjoint subgraphs.

\item The graphical calculus generalises to domains other than just
  vector spaces. In particular, it provides a representation for
  compact closed cartegories.

\end{itemize}

The major problem with these graphical representations is the lack of
machinery for automating the manipulation of graphs.  Existing
approaches to graph transformation have a different underlying
semantics corresponding to graphs in general rather than those which
describe compact closed categories. The result is that they provide an
unsound concept of rewriting for the kind of graphs we are interested
in.

This paper reviews a formalism for graphical representations of
compact closed categories. We also revisit a model of quantum
computation based on this calculus. We then extend the graphical
calculus in two significant ways driven by the need to express rules
that could not be accounted for in the initial formalism. We provide a
semantics for these in terms of the initial representation. By
suitably combining these extensions to graphs, we provide a
representation that captures an interesting and useful set of graph
patterns. We use this as the foundation for a simple logical framework
for manipulating compact closed categories as graphs. This has a
suitable rewriting mechanism where the axioms of the underlying
object-formalism are expressed as equations between graphs. We then
present a short case study that illustrates the framework by
instantiating it for the introduced model of quantum computation.
This shows how the framework can be used to symbolically perform
simplifications to quantum programs as well as simulate quantum
computations.

@@

%Can formalise quantum mechanics in an abstract setting using compact
%or $\dag$-compact categories.

%We will give a description of graphs in general;  a specialisation of
%this gives the structure required for compact closed  categories.  The
%graph structure suffices to give a tight representation.

%We then introduce a concrete set of generators and equations for use
%in quantum computation.  We need rewrites to capture the non-logical
%axioms of this structure.   It turns out that we can most easily express
%these equations using an infinite family of ``spider rules'';  this
%motivates the  more general notion of graph patterns.

@@

\noindent The intuition behind this definition is that the subgraph of
$G$ determined by $V'$ should be preserved exactly by $f$; it should
contain $G$ without any extra incident egdes into the verticies $V'$
whereas other vertices may be identified and may contain additional
verticies.

@@

\begin{theorem}
  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. 
\end{theorem}
